---
title: "Data Cleaning and Configuration"
author: "Jiaheng Cai"
date: "11/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

System information
```{r}
library(tidyverse)
library(readxl)
library(gtsummary)
library(survival)
library(survminer)

os <- sessionInfo()$running

if (str_detect(os, "Ubuntu")) {
  path <- '~/Biostatistics_M215_Project/data/'
} 
if(str_detect(os, "Windows")){
  path <- "C:/Users/alexc/Desktop/215/Biostats-M215/"
}
if(str_detect(os, "mac")){
  path <- "~/Downloads/Biostat 215 Project/Biostatistics_M215_Project/Data/"
}
```

### Data Cleaning and Configuration
The idea is that, at the endpoint dataset, we have 3088 observations. However, in our
covariate dataset, we only have 2333 observations. There is a 755 gap and it is hard
to perform imputation since it is going to be computationally intensive and the result 
will be inaccurate. Thus, we will perform a complete dataset study. We will keep the 
subject which they finished the NDS food study. 

```{r}
endpoint.data = read_excel(paste0(path, 'endpoints.xls'))
ndsfoody4.data = read_excel(paste0(path, 'ndsfoody4.xls'))
colnames(ndsfoody4.data)[1] = 'id'
full.data = merge(endpoint.data, ndsfoody4.data, by.y='id', all.x=TRUE)
full.data = na.omit(full.data) 
```

```{r}
full.data$vitality = as.factor(full.data$vitality)
full.data$recur_flag = as.factor(full.data$recur_flag)

full.data %>%
  tbl_summary(by = 'intgrp',
              missing_text = "Missing",
              include = -c(id, Year_Dx, grade2, grade3, grademiss, stage2, stage3,
                           `last date`)) %>%
  add_p()
```
As one can see, at the endpoint, there is no significant difference between the time
to recurrence and time to death caused by breast cancer from these 2 groups. 

```{r}
full.data$endtimer = full.data$Year_Dx + full.data$yrsdx_endr
# (endpoint.km.fit <- survfit(Surv(yrsdx_endr, recur_flag) ~ 1, data = full.data))
full.data = full.data[order(full.data$Year_Dx),]
rownames(full.data) = NULL
(cbind(full.data$Year_Dx, full.data$endtimer, full.data$recur_flag, full.data$vitality))
```

```{r}
endpoint.data = read_excel(paste0(path, 'endpoints.xls'))
ndsfoody1.data = read_excel(paste0(path, 'ndsfoody1.xls'))
ndsfoody4.data = read_excel(paste0(path, 'ndsfoody4.xls'))

colnames(ndsfoody1.data)[1] = 'id'
colnames(ndsfoody4.data)[1] = 'id'

colnames.4 = colnames(ndsfoody4.data)
for (i in 1:84){
  colnames.4[i] = paste0(colnames.4[i], '4')
}
colnames(ndsfoody4.data) = colnames.4
colnames(ndsfoody4.data)[1] = 'id'

full.data = merge(endpoint.data, ndsfoody1.data, by.y='id', all.x=TRUE)
full.data = merge(full.data, ndsfoody4.data, by.y='id', all.x=TRUE)
full.data = na.omit(full.data) 

full.data$vitality = as.factor(full.data$vitality)
full.data$recur_flag = as.factor(full.data$recur_flag)

full.data %>%
  tbl_summary(by = 'intgrp',
              missing_text = "Missing",
              include = -c(id, Year_Dx, grade2, grade3, grademiss, stage2, stage3,
                           `last date`)) %>%
  add_p()
```

